---
title: "Reproducible Research - Peer Assessment 1"
author: "Stephen Lee"
date: "November 15, 2015"
output: html_document
---

###Part 1: Loading and Preprocessing the data
Before performing any computations, let's load the "Activity Monitoring" dataset and get a feel for the data.

```{r}
activity <- read.csv ("activity.csv")
str (activity)
```



###Part 2: What is the mean total number of steps taken per day?
After loading the dataset, we can now calculate the total number of steps taken per day and plot a histogram showing this.

```{r}
total_steps_daily <- aggregate (steps ~ date, activity, sum, na.rm = TRUE)
head (total_steps_daily)
hist (total_steps_daily$steps, col = "red", main = "Frequency vs. Daily Steps", xlab = "Total  Number of Steps Taken Each Day")
```

Based on the histogram above, the 10,000-15,000 step range has the highest frequency.

Additionally, we can also increase the number of bins in the histogram in order to get a more revealing distribution and better understand our data.

```{r}
histinfo <- hist (total_steps_daily$steps, col = "red", main = "Frequency vs Daily Steps", xlab = "Total  Number of Steps Taken Each Day", breaks = 11, axes = FALSE)
axis (side = 1, at = c (0, 2000, 4000, 6000, 8000, 10000, 12000, 14000, 16000, 18000, 20000, 22000), cex.axis = 0.70)
axis (side = 2, at = seq (0, 16, by = 4))
histinfo
```

Looking at the second histogram now, the 10-12000 step range has the highest count.

Let's now calculate the mean and median total number of steps taken per day. 

```{r}
summary (total_steps_daily)
```

The median and mean total number of steps taken daily are 10765 and 10766, respectively. 


###Part 3: What is the average daily activity pattern?
The second part of this assignment is to create a time series plot of all the 5-minute intervals and the corresponding average number of steps taken, averaged across all days. First, let's compute the average number of steps taken for each 5-minute interval.

```{r}
int_avg <- aggregate (activity$steps, list (activity$interval), mean, na.rm = TRUE)
colnames (int_avg) <- c ("Interval", "Avg_Steps")
head (int_avg)
```

Now, let's create the time series plot.

```{r}
plot (int_avg$Interval, int_avg$Avg_Steps, main = "Avg Steps Vs. Interval", xlab = "Interval", ylab = "Avg Steps Per Interval", type = "l")
```

The y-axis in the plot above shows the average number of steps for each 5-minute interval, whereas the x-axis lists the intervals. The final time interval of each day is 2355, or 23:55/11:55 p.m.

From the time series plot, we can determine the 5-minute interval which has the highest or maximum number of steps.

```{r}
subset (int_avg, Avg_Steps == max (Avg_Steps))
```

The interval in question is 8:35 a.m.



###Part 4: Imputing missing values
First, let's count the number of "NA's" in the "Activity" data set.

```{r}
summary(activity$steps)
```

The total number of missing values is 2304.

Next, let's fill in all the missing values in the dataset in order to reduce the bias that may have potentially crept in when the mean and median of the original data set were computed. Let's replace all the NA's with the mean for each corresponding 5-minute interval.

```{r}
library (plyr)
impute_mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))
activity_new <-ddply (activity, "interval", transform, steps = impute_mean (steps))
activity_new <- activity_new [order(activity_new$date), ]
head (activity_new)
```

Using the newly created dataset, let's repeat the first part of the assigment and calculate the total number of steps taken per day. Next, plot a histogram showing the new data set and compare it with the first histogram.

```{r}
new_total_daily <- tapply (activity_new$steps, activity_new$date, sum)
head (new_total_daily)

hist (total_steps_daily$steps, col = "red", main = "Frequency vs. Daily Steps (Before Imputing Interval Mean)", xlab = "Total  Number of Steps Taken Each Day")
hist (new_total_daily, col = "lightblue", main = "Frequency vs. Daily Steps (After Imputing Interval Mean)", xlab = "Total  Number of Steps Taken Each Day")
```

In the first part of the assignment, I created a second histogram using additional bins. Let's make a comparison for this too.

```{r}
hist (total_steps_daily$steps, col = "red", main = "Frequency vs Daily Steps (Before Imputing Interval Mean)", xlab = "Total  Number of Steps Taken Each Day", breaks = 11, axes = FALSE)
axis (side = 1, at = c (0, 2000, 4000, 6000, 8000, 10000, 12000, 14000, 16000, 18000, 20000, 22000), cex.axis = 0.70)
axis (side = 2, at = seq (0, 16, by = 4))

histinfo_2 <- hist (new_total_daily, col = "lightblue", main = "Frequency vs Daily Steps (After Imputing Interval Mean)", xlab = "Total  Number of Steps Taken Each Day", breaks = 11, axes = FALSE)
axis (side = 1, at = c (0, 2000, 4000, 6000, 8000, 10000, 12000, 14000, 16000, 18000, 20000, 22000), cex.axis = 0.70)
axis (side = 2, at = seq (0, 24, by = 4))
histinfo_2
```

For the first histogram, the 10,000-15,000 step range has the highest frequency, just like its counterpart in part 1. This is also true for the second histogram, where the 10-12000 step range has the highest count.

Let's now calculate the new mean and median total number of steps taken per day. 

```{r}
summary(new_total_daily)
```

The new mean and median are both 10770, which are both very similiar to the first mean (10766) and median (10765). The impact of imputing missing data with the mean estimates is evident in both new histograms, where the distribution is more concentrated around the mean and median of the sample data.



###Part 5: Are there differences in activity patterns between weekdays and weekends?
First, let's create a new factor variable in the dataset with 2 levels - 'weekday' and 'weekend', indicating whether a given date is a weekday or weekend day.

```{r}
wkday <- c ("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
Day_Type <- factor ((weekdays (as.Date (activity_new$date)) %in% wkday), levels = c (TRUE, FALSE), labels = c ("weekday", "weekend"))
final_activity <- as.data.frame (cbind (activity_new, Day = weekdays (as.Date (activity_new$date)), Day_Type))
head (final_activity)
```

Using the new dataset, let's create a time series panel plot of the average number of steps taken for each 5-minute interval, averaged across all weekday and weekend days.

```{r}
int_wkday <- subset (final_activity, Day_Type == "weekday")
int_wkend <- subset (final_activity, Day_Type == "weekend")

int_wkday_avg <- aggregate (int_wkday$steps, list (int_wkday$interval), mean)
colnames (int_wkday_avg) <- c ("Interval", "Avg_Steps")
int_wkend_avg <- aggregate (int_wkend$steps, list (int_wkend$interval), mean)
colnames (int_wkend_avg) <- c ("Interval", "Avg_Steps")

int_wkday_df <- cbind (int_wkday_avg, Day_Type = "weekday")
int_wkend_df <- cbind (int_wkend_avg, Day_Type = "weekend")
int_final <- rbind (int_wkday_df, int_wkend_df)

library (lattice)
int_final <- transform (int_final, Day_Type = factor (Day_Type))
head (int_final)
xyplot(Avg_Steps ~ Interval | Day_Type, int_final, ylab = "Avg Steps Per Interval", type = "l", layout = c (1, 2))
```

Looking at the panel plot, the user is more active during the 8:00-9:30 am time period on weekdays. However, more activity (steps) is oberserved during the 12:30-17:30 time period during the weekends.

